---
title: "Plotting with GGPLOT2"
author: "Tim Dennis"
date: "January 8, 2016"
output: html_document
---
```{r, include=FALSE}
source("tools/chunk-options.R")
opts_chunk$set(fig.path = "fig/08-plot-ggplot2-")
# Silently load in the data so the rest of the lesson works
#gapminder <- read.csv("data/gapminder-FiveYearData.csv", header=TRUE)
```
```{r}
#install.packages('ggplot2')
library(ggplot2)
```


GOALS: Students should be able to use ggplot2 to generate publication quality graphics and understand and use the basics of the grammar of graphics. 

* GGPLOT2 developed by Hadley Wickham based on a *grammar-of-graphics*
* Grammar of graphics consists of a **dataset**, **coordinate system**, and **geoms** -- the visual representation of data points. 
* Think about figure in layers:  like you would in photoshop, illustrator or inkscape.


```{r lifeExp-vs-gdpPercap-scatter, message=FALSE}
library(ggplot2)
gapminder <- read.csv("https://goo.gl/BtBnPg", header = T)
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
  geom_point()
```

NOTE: 

* First we call the `ggplot` function -any arguments we provide the `ggplot` function are considered global options: they apply to all layers
* We passed two arguments to `ggplot`: 
  * data 
  * an `aes` function - which tells ggplot how variables map to aesthetic properties 
  * `x` & `y` locations 
  
Alone the ggplot call isn't enough to redner the plot. 

```{r no_geom, eval=FALSE}
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap))
## If run, would produce an error. 
```

Need to tell ggplot how we want to present variables by specifying a geom layer. In the above example we used `geom_point` to create a scatter plot. 

```{r lifeExp-vs-gdpPercap-scatter2}
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
  geom_point()
```

## Layers

Using scatter plot not the best way to visualize change over time. Let's use line plot. 

```{r}
ggplot(data = gapminder, aes(x=year, y=lifeExp, by=country, color=continent)) +
  geom_line()
```

* used geom_line instead of geom_point 
* added a **by** *aesthetic* to get a line per country and color by contintent 


* visualise both lines and points on the plot? 

```{r lifeExp-line-point}
ggplot(data = gapminder, aes(x=year, y=lifeExp, by=country, color=continent)) +
  geom_line() + geom_point()
```

* note this is layered: so points have been drawn on top of the lines. 
* example of this

```{r lifeExp-layer-example-1}
ggplot(data = gapminder, aes(x=year, y=lifeExp, by=country)) +
  geom_line(aes(color=continent)) + geom_point()
```

* the *aesthetic* mapping of **color** has been moved from the
global plot options in `ggplot` to the `geom_line` layer so it no longer applies to the points 

## Transformations and statistics

* overlay statistical models over the data
* let's use our first example

```{r lifeExp-vs-gdpPercap-scatter3, message=FALSE}
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap, color=continent)) +
  geom_point()
```

* hard to see relationships b/t points because of strong outliers in GDP/cap
* change the scale of units on the y axis using the *scale* functions (controls the mapping between the data values and
visual values of an aesthetic)

```{r axis-scale}
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
  geom_point() + scale_y_log10()
```

* `log10` function applied a transformation to the values of the gdpPercap
column before rendering them on the plot
* each multiple of 10 now only corresponds to an increase in 1 on the transformed scale, e.g. a GDP per capita
of 1,000 is now 3 on the y axis, a value of 10,000 corresponds to 4 on the y
axis and so on
* This makes it easier to visualise the spread of data on the y-axis.

* We can fit a simple relationship to the data by adding another layer, `geom_smooth`:

```{r lm-fit}
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
  geom_point() + scale_y_log10() + geom_smooth(method="lm")
```

*  make the line thicker by *setting* the **size** aesthetic in the `geom_smooth` layer:

```{r lm-fit2}
pwd <- ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
  geom_point() + scale_y_log10() + geom_smooth(method="lm", size=1.5)
```

* two ways an *aesthetic* can be specified:
  1. Here we *set* the **size** aesthetic by passing it as an argument to `geom_smooth`. 
  2. use the `aes` function to define a *mapping* between data variables and their visual representation.

## Multi-panel figures

* we can split this out over multiple panels by adding a layer of **facet** panels:

```{r facet}
ggplot(data = gapminder, aes(x = year, y = lifeExp, color=continent)) +
  geom_line() + facet_wrap( ~ country)
```

* `facet_wrap` layer took a "formula" as its argument, denoted by the tilde
(~). 
* tells R to draw a panel for each unique value in the country column
of the gapminder dataset.

## Modifying text

* would like to add text to elements in the graph
* do this by adding a few more layers:  
   * **theme** layer controls axis text & text size
   * **scales** layer to change legend title

```{r theme}
ggplot(data = gapminder, aes(x = year, y = lifeExp, color=continent)) +
  geom_line() + facet_wrap( ~ country) +
  xlab("Year") + ylab("Life expectancy") + ggtitle("Figure 1") +
  scale_fill_discrete(name="Continent") +
  theme(axis.text.x=element_blank(), axis.ticks.x=element_blank())
```

## Resources: 

This is just a taste of what you can do with `ggplot2`. RStudio provides a
really useful [cheat sheet][cheat] of the different layers available, and more
extensive documentation is available on the [ggplot2 website][ggplot-doc].
Finally, if you have no idea how to change something, a quick google search will
usually send you to a relevant question and answer on Stack Overflow with reusable
code to modify!

[cheat]: http://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf
[ggplot-doc]: http://docs.ggplot2.org/current/