---
title: "Exploring Data Frames"
author: "Tim Dennis"
date: "11/7/2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**How can I manipulate a dataframe?**

* At this point, you’ve see it all - in the last lesson, we toured all the basic data types and data structures in R. Everything you do will be a manipulation of those tools. 
* star of show == data.frame
* we learned that data.frames were vecotrs, so that we have consistent data down a column
* if we want a new column we start by making a vector

```{r}
cats <- read.csv(file="data/feline-data.csv")
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
age <- c(2,3,5,12)
cats
```

* we can then add a column via: 

```{r eval=FALSE}
cats <- cbind(cats, age)
```
* what happened? 
*  Of course, R wants to see one element in our new column for every row in the table:

```{r}
cats
age <- c(4,5,8)
cats <- cbind(cats, age)
cats
```

Now how about adding rows - in this case, we saw last time that the rows of a data.frame are made of lists:

```{r}
newRow <- list("tortoiseshell", 3.3, TRUE, 9)
cats <- rbind(cats, newRow)
```

* another thing - when R creates a factor, it only allows whatever is originally there when our data was loaded ('black', 'calico', 'tabby')
* anything that doesn't fit into one of its categories is rejected
* we need to explicitly add it as a level

```{r}
levels(cats$coat)
levels(cats$coat) <- c(levels(cats$coat), 'tortoiseshell')
cats <- rbind(cats, list("tortoiseshell", 3.3, TRUE, 9))
```

* Alternatively we can change a factor column to a character vector
* but we lose handy categories of the factor but can subseqently add any word we want to the column 

```{r}
str(cats)
cats$coat <- as.character(cats$coat)
str(cats)
```

* we now know how to add rows and columns to our data.fram in R 
* but in our work we've accidentlly added a garbage row

```{r}
cats
```
* we can ask for a data.fram minus this offending row: 
```{r}
cats[-4,]
```

* notice the comma with nothing after it to indicate we want to drop that row
* **Note** we can remove both new rows at once by putting the row numbers inside of a vector: `cats[c(-4,-5)]
* we can also drop all rows with NA values

```{r}
na.omit(cats)
```

* let's reassign the output to cats, so our canges are made permanent

```{r}
cats <- na.omit(cats)
```
* the key to remember when adding data to a data.frame is that columns are **vectors or factors, and rows are lists**
* we can also glue two data.frames together with rbind:

```{r}
cats <- rbind(cats, cats)
cats
```

* but now row names are complicated 
* we can remove the rownames and R will auto re-name them sequentially

```{r}
rownames(cats) <- NULL
cats
```

## Challenge 1
<http://swcarpentry.github.io/r-novice-gapminder/05-data-structures-part2/#challenge-1>

**So far:**

* basics of manipulating `data.frames` with our cat data
* let's work on a real data set

```{r}
gapminder <- read.csv("data/gapminder-FiveYearData.csv")
```

* mention tab-separated values files (.tsv)
* to specify a tab separator use "\\tab" or read.delim()
* files can be downloading via the internet using the `download.file` and the read.csv can be executed to read the downloaded file such as: 

```{r}
download.file("https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/gh-pages/_episodes_rmd/data/gapminder-FiveYearData.csv", destfile = "data/gapminder-FiveYearData.csv")
gapminder <- read.csv("data/gapminder-FiveYearData.csv")
```

* alternately you can read files directly into R from the internet by replacing file paths with a web address in read.csv
* **note** when you do this there is not **local copy** saved

```{r}
gapminder <- read.csv("https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/gh-pages/_episodes_rmd/data/gapminder-FiveYearData.csv")
```

* You can read directly from excel spreadsheets without converting them to plain text by using the readxl package 
* let's investigate gapminder
```{r}
str(gapminder)
```
```{r}
typeof(gapminder$year)
typeof(gapminder$lifeExp)
typeof(gapminder$country)
str(gapminder$country)

```

* we can also interrogat the data.frame for info about itself
* remember that str(gapminder) said there were 1704 observations of 6 variables in gapminder
* what do you think the following will produce?

```{r}
length(gapminder)
```

* a fair guess would say that the length of a data.frame would be the number of rows it has (1704)
* not the case, remember, a data.frame is a *list of vectors and factors*:

```{r}
typeof(gapminder)
```

* when length gave us 6 it's because gapminder is built out of a list of 6 columns 
* to get number of rows and columns in our data set try:

```{r}
nrow(gapminder)
ncol(gapminder)
```

* or both at once

```{r}
dim(gapminder)
```

* we also would want to know what the titles of all the columns are, so we can ask for them by name:

```{r}
colnames(gapminder)
```
* it is important to ask if the structure R is reporting matches our intuition or expectations
* do the basic dat a types reported make sense?
* if not we need to sort out problems now before they turn into negative surprises down the road

* once we are happy that the data types and structures seem reasonable, it's time to start digging into our data properly

```{r}
head(gapminder)
```
* to make sure out analysis is reproducible we should put the code into a script file so we can come back later


## Challenge 2
<http://swcarpentry.github.io/r-novice-gapminder/05-data-structures-part2/#challenge-2>

## Challege 3
<http://swcarpentry.github.io/r-novice-gapminder/05-data-structures-part2/#challenge-3>






