---
title: "Subsetting Data"
author: "Tim Dennis"
date: "11/7/2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

* To be able to subset vectors, factors, matrices, lists, and data frames
* To be able to extract individual and multiple elements: by index, by name, using comparison operations
* To be able to skip and remove elements from various data structures.

## Subsetting Data

* Six different ways we can subset any kind of object
* Three different subsetting operators for the different data structures
* let's start with vectors

```{r}
x <- c(5.4, 6.2, 7.1, 4.8, 7.5)
names(x) <- c('a', 'b', 'c', 'd', 'e')
x
```

## Accessing elements using their indices

* extract elements of a vector we can give their corresp. indes
```{r}
x[1]
x[4]
```

* it may not look like it but the square brackets operator is a function
* it means: "get me the nth element"
* ask for multiple elements at once

```{r}
x[c(1, 3)]
```

* or slices of the vector

```{r}
x[1:4]
```

* the `:` operator creates a sequence of numbers from the left element to the right

```{r}
1:4
c(1, 2, 3, 4)
```

* we can ask for the same element multiple times

```{r}
x[c(1,1,3)]
```
* if we ask for a number outside the vecotr, R will return missing values:

```{r}
x[6]
```

* this is a vector of length of one containing an NA, whose name is also NA
* if we ask for the 0th element, we get an empty vector:

```{r}
x[0]
```

* R indexing starts at 1 - many programming languages (c & python) indexing starts at 0

## Skipping and removing elements

* if we use a negative number as the index of a vector, R will return every element EXCEPT for the one specified

```{r}
x[-2]
```

* we can skip multiple elements

```{r}
x[c(-1,-5)] #or x[-c(1,5)]
```
* a common trip up for novices occurs when trying to skip slices of a vector
* most ppl first try to negate a sequence like so:

```{r eval=FALSE}
x[-1:3]
```
cryptic error: 

```{r eval=FALSE}
Error in x[-1:3]: only 0's may be mixed with negative subscripts
```

* remember order of operations. 
* : is really a function
  * what happens is it takes first arguemnt as -1, and a second as 3, so getnerates a sequence of numbers: c(-1, 0, 2, 3)

* correct solution is to wrap that function call in brackets, so that the - operator applies tot he result

```{r}
x[-(1:3)]
```

* to remove elements from a vector we need to assign the results back onto the object

```{r}
x <- x[-4]
x
```
## challenge 1

<http://swcarpentry.github.io/r-novice-gapminder/06-data-subsetting##challenge-1>

## Subsetting by name

* extract elements by using their name
* `x[c("a","c")]` 
    - usuall more reliable way to subset objects, index changes more than names
    - unfort can't skip or remove as easily
```{r}
x[-which(names(x)=="a")]
```

* which function returns all true elements of its argument, which is negated 
* let's step through this: 

```{r}
names(x) == "a" # the condition applied to vector x, only a true
which(names(x)=="a") # which converts this to a index
x[-which(names(x)=="a")] # using the "-" will remove the element at that index
```

* skipping multiple named indicies uses %in%

* `x[-which(names(x) %in% c("a", "c"))]`
* So why can’t we use == like above? 
* Because '==' expects objects of the same length and if not, will start repeating the shorter object
* Called recycling rule 

## challenge 2

<http://swcarpentry.github.io/r-novice-gapminder/06-data-subsetting##challenge-1>
**R recycling rule** 

```{r eval=F}
names(x) == c('a', 'c') #warnings 
#== works slightly differently than %in%. It will compare each element of its left argument to the corresponding element of its right argument.
# R recycles the shorter vector in a equality comparison

c("a", "b", "c", "e")  # names of x
   |    |    |    |    # The elements == is comparing
c("a", "c")

c("a", "b", "c", "e")  # names of x
   |    |    |    |    
c("a", "c", "a", "c") # The elements == is comparing
```

*  Non-unique names are allowable for multiple elements in a vector 

```{r}
x <- 1:3
x
names(x) <- c('a', 'a', 'a')  
x
x['a']  # only returns first value
x[which(names(x) == 'a')]  # returns all three values
```

* Getting help for operators; 
    - `help("%in%")` or `?"%in%"`

**Subsetting through other logical operators**

* we can subset through logical operations
* `x[c(TRUE, TRUE, FALSE, FALSE)]`
* Note that in this case, the logical vector is also recycled to the length of the vector we’re subsetting!
* `x[c(TRUE, FALSE)]`
* Since comparison operators evaluate to logical vectors, we can also use them to succinctly subset vectors:
    - `x[x > 7]`

There are many situations in which you will wish to combine multiple conditions. To do so several logical operations exist in R:

* | logical OR: returns TRUE, if either the left or right are TRUE.
* & logical AND: returns TRUE if both the left and right are TRUE
* ! logical NOT: converts TRUE to FALSE and FALSE to TRUE
* && and || compare the individual elements of two vectors. Recycling rules also apply here.

**Handling special values**

At some point you will encounter functions in R which cannot handle missing, infinite, or undefined data.

special functions to deal with this: 

* is.na will return all positions in a vector, matrix, or data.frame containing NA.
* likewise, is.nan, and is.infinite will do the same for NaN and Inf. 
* is.finite will return all positions in a vector, matrix, or data.frame that do not contain NA, NaN or Inf.
* na.omit will filter out all missing values from a vector

**factor subsetting**

Factor subsetting works the same way as vector subsetting.

```{r}
f <- factor(c("a", "a", "b", "c", "c", "d"))
f[f == "a"]
f[f %in% c("b", "c")]
f[1:3]

```
* An important note is that skipping elements will not remove the level even if no more of that category exists in the factor:
    - f[-3]

**Matrix subsetting**

Matrices are also subsetted using the [ function. In this case it takes two arguments: the first applying to the rows, the second to its columns:

```{r}
set.seed(1)
m <- matrix(rnorm(6*4), ncol=4, nrow=6)
m[3:4, c(3,1)]
```
* You can leave the first or second arguments blank to retrieve all the rows or columns respectively:

```{r}
m[, c(3,4)] #column
m[3,] #if grabbing 1 row, R will convert to vector
m[3, , drop=FALSE] # keep as a matrix specify a third argument; drop = FALSE:
#m[, c(3,6)] #will throw error if out of range
```

* because matricies are really vectors, you can just use single indexing
    - m[5] #not very useful
* matricies are populated by column-major format by default, elements are arraged column wise

```{r}
matrix(1:6, nrow=2, ncol=3)
matrix(1:6, nrow=2, ncol=3, byrow=TRUE) #populate by row use  byrow=TRUE

```
## challenge 4

<http://swcarpentry.github.io/r-novice-gapminder/06-data-subsetting#challenge-4>

## List subsetting

Three functions to subset lists [, [[, $

* Using [ will always return a list. If you want to subset a list, but not extract an element, then you will likely use [.

```{r}
xlist <- list(a = "Software Carpentry", b = 1:10, data = head(iris))
xlist[1]
```

* we can subset elements teh same way as atomic vector
* xlist[1:2]
* to get at individual elements of a list you need to use [[
    - `xlist[[1]]`
    - but you cant extract more than one element at once `xlist[[1:2]]`
    - nor skip elements `xlist[[-1]]`
    - but you can use names to subset and extract `xlist[['a']]`
    - $ function extracts by name xlist$data

## challenge 5

<http://swcarpentry.github.io/r-novice-gapminder/06-data-subsetting##challenge-5>

## Data frame

* Data frames are lists under the hood, so similar methods apply.
* [ with one argument will act the same was as for lists, where each list element corresponds to a column. The resulting object will be a data frame:

```{r}
gapminder <- read.csv(file="data/gapminder-FiveYearData.csv")
head(gapminder[3])
head(gapminder[["lifeExp"]]) # [[ will act to extract a single column
head(gapminder$year) #$ provides shorthand to extract columns by name
#With two arguments, [ behaves the same way as for matrices:
gapminder[1:3,]
gapminder[3,] #is a data frame b/c of the mixed types
```
## Challenge 7 

<http://swcarpentry.github.io/r-novice-gapminder/06-data-subsetting##challenge-7>








