---
title: "Vectorization"
author: "Tim Dennis"
date: "11/7/2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

* most of R's functions are vectorized -- meaning the function will operate on all elements of a vector without needing to loop throu and act on each element one at a time
* this makes writing code more concise, easy to read, and less error prone

```{r}
x <- 1:4
x * 2
```

* multiplication above happened to each element of a vector
* we can also add two vectors together

```{r}
y <- 6:9
x + y
```
something liek this:

```{r eval=FALSE}
x:  1  2  3  4
    +  +  +  +
y:  6  7  8  9
---------------
    7  9 11 13
```

## Challenge 1 

<http://swcarpentry.github.io/r-novice-gapminder/09-vectorization#challenge-1>

* comparison operators, logical operators and functions are vectorized

**comparison operators**

```{r}
x
x > 2
```

* any() will return TRUE if any element of a vector is TRUE all() will return TRUE if all elements of a vector are TRUE

* most functions also operate element-wise on vectors

**functions**

```{r}
x <- 1:4
log(x)
```
* vectorized operatos work element-wise on matrices

```{r}
m <- matrix(1:12, nrow=3, ncol=4)
m * -1
```



