---
title: "Homework 2 - R GGPLOT2"
author: "Tim Dennis & Reid Otsuji"
date: "1/21/2017"
output: 
  html_document: 
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Due 2/4 - By Midnight 

### How to submit your homework:
1. Download the [r-homework2.R](r-homework2.R) script and open in Rstudio
2. Rename `r-homework2.R` to `yourlastname_sudentID_r-homework1.R`
3. Type the answers in the uncommented lines (those without #). Notice, for some lines, we have gotten you started with names of objects.
4. Save the script file.
5. Upload the file to tritonEd (we are working on getting that set up).  


**Remeber to load the ggplot2 package and read in gapminder data**

```{r}
setwd('~/workshops/data-workshops/notebooks/intro-r')
gapminder <- read.csv('data/gapminder-FiveYearData.csv')
library(ggplot2)
```

## Question 1. Adding text to the graph

The following code produces multiple panels of life expectancy over time from `gapminder` focusing only on those countries with names that start with the letter “A” or “Z”.

We start by subsetting the data. We use the substr function to pull out a part of a character string; in this case, the letters that occur in positions start through stop, inclusive, of the gapminder$country vector. The operator %in% allows us to make multiple comparisons rather than write out long subsetting conditions (in this case, starts.with %in% c("A", "Z") is equivalent to starts.with == "A" | starts.with == "Z"). Run the code to get a feel for the steps involved. 

```{r}
starts.with <- substr(gapminder$country, start = 1, stop = 1)
az.countries <- gapminder[starts.with %in% c("A", "Z"), ]
ggplot(data = az.countries, aes(x = year, y = lifeExp, color=continent)) +
  geom_line() + facet_wrap( ~ country)
```

To clean this figure up for a publication we need to change some of the text elements. The x-axis is too cluttered, and the y axis should read “Life expectancy”, rather than the column name in the data frame.

We can do this by adding a couple of different layers. The theme layer controls the axis text, and overall text size, and there are special layers for changing the axis labels. To change the legend title, we need to use the scales layer. 

Set the title to "Figure 1", Y label to "Year", X Label to "Life expectancy", and Legend title to "Continent". Use these functions to set those. 

* title = ggtitle()
* X label = xlab()
* Y label = ylab()
* Legend title = scale_colour_discrete()

ANSWER: 
```{r eval=FALSE}
starts.with <- substr(gapminder$country, start = 1, stop = 1)
az.countries <- gapminder[starts.with %in% c("A", "Z"), ]
ggplot(data = az.countries, aes(x = year, y = lifeExp, color=continent)) +
  geom_line() + facet_wrap( ~ country)
## your added functions to set the title, xlab, ylab and legend here
```


## Question 2. Basic scatter plot

Use the **mtcars** dataset available in RStudio (look at the help: `?mtcars`).  The data was extracted from the 1974 Motor Trend US magazine, and comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973–74 models).   Create a scatter plot that shows the relationship between various car weights (wt), miles per gallon (mpg) and engine cylinders (cyl). Use colored points to show the different cylinders in the plot. Note: you will need to convert cyl to a factor. You will need the function `factor()` to do this. 

ANSWER:
```{r echo=FALSE}
ggplot(mtcars, aes(wt, mpg)) + 
  geom_point(aes(colour = factor(cyl)))
```

## Question 3. Adding shapes

Using the solution from the previous example, create a new plot using shapes to differentiate the various engine cylinders.  

ANSWER:
```{r echo=FALSE}
ggplot(mtcars, aes(wt, mpg)) + 
  geom_point(aes(shape = factor(cyl), color = factor(cyl)))
```

## Question 4. Assign a ggplot function to an object.

Create a ggplot function that contains the built-in `diamonds` dataset and an aesthetics layer.  Map `carat` on the x axis and `price` on the y axis.  Assign the ggplot function to an object: **myplot**

Using **+** , add a `geom_point` with an alpha of 0.1. Then add (`+`) geom_smooth and `theme_minimal` layers to the `myplot` object.  Map **clarity** to the color argument in the `geom_smooth` function (add **se=FALSE** to `geom_smooth` to turn off error shading). 
 
Note: the plot will take a few seconds to generate. 

ANSWER:
```{r echo=FALSE}
Myplot <- ggplot(data = diamonds, aes(carat, price))
Myplot + geom_point(alpha=0.1) + geom_smooth(aes(color=clarity), se=FALSE) + theme_gray()
```


## Question 5. Creating an area graph

Using the gapminder data we used on Wed.’s class, create a density plot of GDP per capita, filled by continent
that looks like the following: 
NOTE: Transform the x axis using a log10 to better visualise the data spread. - Add a facet layer to panel the density plots by year. 

Hint: Look up the `facet_wrap` and `scale_x_*` functions to help you accomplish this. 

ANSWER:
```{r echo=FALSE}
ggplot(data = gapminder, aes(x = gdpPercap, fill=continent)) +
 geom_density(alpha=0.6) + facet_wrap( ~ year) + scale_x_log10()
```